---
title: "STA 141A:  Predicting Wildfire Incidence in the United States"
author: "(Zhuyue Chen: 923485896)"
output: html_document
---

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message = FALSE,warning = FALSE)
```





# 1 Abstract

This project develops a predictive framework for identifying large wildfires across the United States. Wildfire incidents are aggregated into a spatio-temporal panel dataset defined at the $1^\circ \times 1^\circ$ grid-cell level with monthly resolution from $1992$ to $2015$. The prediction task is formulated as a binary classification problem that forecasts whether a wildfire of at least 300 acres will occur in a grid cell in the following month. 

Two predictive models are evaluated: logistic regression as an interpretable statistical baseline and XGBoost as a nonlinear machine learning model. Model performance is assessed using time-aware expanding-window cross-validation to avoid temporal data leakage, with predictive accuracy measured using the Area Under the ROC Curve (AUC). 

The results show that XGBoost consistently outperforms logistic regression, improving average AUC by approximately $12.3\%$. This improvement suggests that nonlinear spatial and temporal patterns play an important role in predicting wildfire occurrence. Incorporating such complex relationships can support more effective early-warning systems and wildfire risk management strategies.
 
 

# 2 Data Preprocessing


In the data preprocessing stage, raw wildfire records are transformed into a structured dataset suitable for predictive modeling. This process includes filtering notable wildfires, constructing spatial–temporal units, defining the binary prediction target, generating predictive features while avoiding time leakage, and handling missing values. The transformation converts individual wildfire observations into a panel dataset in which each row represents a specific geographic grid cell and month.

We begin by loading a sample of 100,000 wildfire records and selecting several key variables that describe wildfire characteristics and location.


| Variable      | Description                                      |
| ------------- | ------------------------------------------------ |
| FIRE_YEAR     | Year when the wildfire was discovered            |
| DISCOVERY_DOY | Day of the year when the wildfire was discovered |
| FIRE_SIZE     | Final burned area of the wildfire (in acres)     |
| LATITUDE      | Latitude of the wildfire ignition location       |
| LONGITUDE     | Longitude of the wildfire ignition location      |
| STATE         | U.S. state where the wildfire occurred           |





```{r}
suppressMessages({
     library(dplyr)
    library(ggplot2)
    library(tidyverse)
    library(slider)
    library(pROC)
    library(xgboost)
    library(Matrix)})


set.seed(2026)
```


```{r}
wildfires<-read_csv(file = "./Data/wildfires_sample_100k.csv")%>%
  select(FIRE_YEAR, DISCOVERY_DOY, FIRE_SIZE, LATITUDE, LONGITUDE, STATE) %>%
  mutate(STATE = as.factor(STATE))

glimpse(wildfires)

```


The variables FIRE_YEAR and DISCOVERY_DOY are converted into a proper date format in order to extract year and month information for temporal analysis.


```{r}
fires <- wildfires %>% 
    mutate(FIRE_DATE = as.Date(paste(FIRE_YEAR, DISCOVERY_DOY, sep = "-"), format = "%Y-%j"),
          YEAR = year(FIRE_DATE),
          MONTH = month(FIRE_DATE))
```


A notable wildfire is defined as a wildfire with a final burned area of at least $300$ acres.
This threshold is consistent with operational wildfire reporting standards used by U.S. fire management agencies, where large wildfires are typically defined as fires exceeding approximately $300$ acres in grass or brush fuel models.


```{r}

notable_threshold_acres <- 300

notable_fires=fires%>%filter(FIRE_SIZE>=notable_threshold_acres)

cat("Number of notable fires:", nrow(notable_fires))
```


To construct spatial units, geographic coordinates are converted into $1^\circ \times 1^\circ$ grid cells by taking the floor of the latitude and longitude values. A year–month identifier is also created to represent the temporal dimension of the data.

```{r}

notable_fires <- notable_fires %>%
  mutate(
    lat_bin = floor(LATITUDE),
    lon_bin = floor(LONGITUDE),
    ym = YEAR * 100 + MONTH   
  )
```



For each grid cell $i$ and month $t$, we create a binary indicator representing whether at least one notable wildfire occurred in that grid cell during that month. The prediction target is defined as a forward-looking variable, indicating whether a notable wildfire occurs in the next month.

$$
y_{i,t+1} = 
\begin{cases} 
1 & \text{if a notable wildfire occurs in grid cell } i \text{ during month } t+1, \\
0 & \text{otherwise}.
\end{cases}
$$
This forward-looking definition ensures that predictions only rely on information available at time $t$, thereby avoiding the use of future information.


```{r}

grid_month_fire <- notable_fires %>%
   group_by(lat_bin,lon_bin,ym)%>%
  summarize(
    n_fires=n(),
    fire_this_month=1L,
    .groups="drop"
  )



head(grid_month_fire)


```



To create a balanced spatial–temporal dataset, we construct a complete panel containing all combinations of grid cells and months observed in the study period. Missing values are filled with zeros.

```{r}

all_grids <- grid_month_fire %>% distinct(lat_bin, lon_bin)

# Generate a safe monthly sequence using actual dates
start_date <- ymd(paste0(min(grid_month_fire$ym) %/% 100, "-", min(grid_month_fire$ym) %% 100, "-01"))
end_date   <- ymd(paste0(max(grid_month_fire$ym) %/% 100, "-", max(grid_month_fire$ym) %% 100, "-01"))

all_months <- tibble(date = seq.Date(start_date, end_date, by = "month")) %>%
  mutate(year = year(date), month = month(date), ym = year * 100 + month) %>%
  select(ym, year, month)

# Full panel = grid × month

panel <- crossing(all_grids, all_months) %>%
  left_join(grid_month_fire, by = c("lat_bin", "lon_bin", "ym")) %>%
  replace_na(list(n_fires = 0L, fire_this_month = 0L))  




glimpse(panel)


```






```{r}

panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    y_next = lead(fire_this_month, n = 1L)
  ) %>%
  ungroup() %>%
  filter(!is.na(y_next))

table(panel$y_next)


```

The dataset contains 1,311 positive observations and 146,265 negative observations, indicating a strong class imbalance. To account for this imbalance, the XGBoost model uses the scale_pos_weight parameter to increase the relative importance of positive wildfire events during model training.





To incorporate geographic information, each grid cell is assigned to a U.S. state based on the most frequently observed state among wildfire records located in that grid.


```{r}



mode_state <- function(x) {
  uniqv <- unique(x)
  uniqv[which.max(tabulate(match(x, uniqv)))]
}





# grid_state <- notable_fires %>% ...
grid_state <- notable_fires %>%
  group_by(lat_bin, lon_bin) %>%
  summarise(
    state = mode_state(STATE),
    .groups = "drop"
  ) %>%
  mutate(state = as.factor(state))

#########

panel <- panel %>%
  left_join(grid_state, by = c("lat_bin","lon_bin"))

panel %>% select('state') %>% summary



```


Seasonal effects are also important in wildfire occurrence. Because months exhibit a cyclical structure (for example, December and January are temporally close despite being numerically distant), we transform the month variable using sine and cosine functions to capture cyclical seasonal patterns.


```{r}

panel <- panel %>%
  mutate(
    month_rad = 2 * pi * month / 12,
    month_sin = sin(month_rad),
    month_cos = cos(month_rad)
  ) %>%
  select(-month_rad)  

#########

panel %>% select('month_sin', 'month_cos') %>% head




```


To account for historical wildfire activity, we construct several lagged and rolling-window predictors based on past fire counts and fire sizes. These features include the number of fires in the previous 12 months, rolling fire counts over three and twelve months, and rolling averages of fire sizes. All rolling statistics are computed using only past information to ensure that no future data is used in the prediction process.



```{r}
 
grid_fire_size <- notable_fires %>%
  group_by(lat_bin, lon_bin, ym) %>%
  summarise(fire_size = mean(FIRE_SIZE, na.rm = TRUE), .groups = "drop")

```


```{r}
panel <- panel %>%
  left_join(grid_fire_size, by = c("lat_bin","lon_bin", "ym")) %>%
  mutate(fire_size = coalesce(fire_size, 0))
```



```{r}

panel <- panel %>%
  arrange(lat_bin, lon_bin, ym) %>%
  group_by(lat_bin, lon_bin) %>%
  mutate(
    fire_lag12  = as.numeric(lag(n_fires, n = 12L, default = 0L)),
    fire_roll3  = slide_dbl(n_fires, sum, .before = 2L, .after = 0L, .complete = FALSE, .default = 0),
    fire_roll12 = slide_dbl(n_fires, sum, .before = 11L, .after = 0L, .complete = FALSE, .default = 0),
     
    size_roll3 = slide_dbl(fire_size, 
                           ~mean(.x[.x > 0], na.rm = TRUE), 
                           .before = 2L, .after = 0L),
    
    
    size_roll12 = slide_dbl(fire_size, 
                            ~mean(.x[.x > 0], na.rm = TRUE), 
                            .before = 11L, .after = 0L)) %>%
  ungroup() %>%
  mutate(across(starts_with("size_roll"), ~replace_na(.x, 0)))
```


Summary of fire history predictors:

```{r}
panel %>% 
  select("n_fires", "fire_lag12", "fire_roll3", "fire_roll12", 
         "size_roll3", "size_roll12") %>%
  summary
```


The summary table is shown below.

| Component               | Description                                           |
|-------------------------|-------------------------------------------------------|
| Rows                    | 147576                                                |
| Time span               | 1992 - 2015                                          |
| Spatial unit            | 1° × 1° latitude–longitude grid                      |
| Temporal unit           | Monthly                                              |
| Outcome variable        | y_next = 1 if a notable fire occurs in the next month|
| Spatial predictors      | lat_bin, lon_bin, state                               |
| Seasonal predictors     | month_cos, month_sin, year                            |
| Fire history predictors | n_fires, fire_lag12, fire_roll3, fire_roll12,<br>size_roll3, size_roll12 |


The detailed description is shown below.

| Component                     | Description                                         |
|-------------------------------|-----------------------------------------------------|
| Outcome variable              | Binary outcome: whether any notable fire occurs in the grid cell next month |
| **Group A — Static**           | latitude: Spatial patterns                           |
|                               | longitude: Spatial patterns                          |
|                               | state: Administrative information                   |
| **Group B — Seasonal**         | Month (cyclical encoding)                            |
|                               | Year                                                |
| **Group C — Lagged fire history** | Fire count this month: Short-term persistence       |
|                               | Fire count lag 12: Seasonal recurrence             |
|                               | Rolling fire counts over past 3 months: Recent intensity |
|                               | Rolling fire counts over past 12 months: Long-term activity |
|                               | Rolling average fire size over past 3 months       |
|                               | Rolling average fire size over past 12 months      |


# 3 Exploratory Data Analysis

## 3.1 Spatial Patterns

To understand the spatial distribution of wildfire risk, we calculate the mean value of the outcome variable $y_next$ for each grid cell. This presents the probability that a notable 



```{r}

eda_spatial <- panel %>%
  group_by(lat_bin, lon_bin) %>%
  summarise(
    fire_rate = mean(y_next, na.rm = TRUE),
    .groups = "drop"
  )



ggplot(eda_spatial, aes(x = lon_bin, y = lat_bin, fill = fire_rate)) +
  coord_quickmap(xlim = c(-125, -65), ylim = c(25, 50)) + 
  geom_tile() +
  labs(
    title = "Average Probability of Notable Fire Next Month by Grid Cell",
    x = "Longitude bin (floor)",
    y = "Latitude bin (floor)",
    fill = "Mean y_next"
  )


```


The spatial distribution reveals clear geographic heterogeneity in wildfire risk across the United States. Higher wildfire probabilities are concentrated in the western and southern regions, while the northeastern region exhibits relatively low wildfire risk.

Some of the highest wildfire probabilities appear in grid cells in the western U.S., particularly in parts of California, Arizona, and Texas, as well as in southern Florida. In contrast, large areas of the northeastern United States show very low wildfire occurrence rates.

These patterns also suggest the presence of spatial autocorrelation, where neighboring grid cells tend to exhibit similar wildfire probabilities. Such spatial clustering indicates that wildfire risk is influenced by geographic factors such as climate conditions, vegetation types, and land use.

The existence of spatial heterogeneity and clustering suggests that the relationship between predictors and wildfire occurrence may be complex and nonlinear. This observation motivates the inclusion of spatial predictors such as latitude, longitude, and state in the predictive models.


## 3.2 Seasonal Patterns

Wildfires also exhibit strong seasonal patterns. To explore this pattern, we compute the average probability of a wildfire occurring in the following month for each calendar month.


```{r}


eda_month <- panel %>%
  group_by(month) %>%
  summarise(
    mean_y_next = mean(y_next, na.rm = TRUE),
    .groups = "drop"
  )


#########

ggplot(eda_month, aes(x = month, y = mean_y_next)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Mean Probability of Next-Month Notable Fires by Season by Month",
    x = "Month",
    y = "Mean of y_next"
  )
```


The results show that wildfire probabilities remain relatively low during winter and early spring. Starting in May, the probability begins to increase rapidly and reaches its peak during the summer months, particularly in July. After this peak, wildfire probability gradually declines toward the autumn months, with the lowest probabilities observed in November.

The peak in July indicates that July conditions strongly predict wildfire risk in August, which corresponds to the height of the wildfire season in many parts of the United States. Similarly, the increase beginning in May signals the transition into the summer wildfire season.

These clear cyclical patterns justify the use of sine and cosine transformations of the month variable. Using a simple linear month variable would incorrectly treat December and January as distant months, whereas the sine–cosine cyclical transformation accurately captures the repeating seasonal structure of wildfire occurrence.

# 4 Modeling


## 1 Baseline model: Logistic Regression

The baseline model used in this study is logistic regression, which is suitable for binary classification problems. 

The model estimates the probability that a notable wildfire occurs in grid cell $i$ during month $t+1$
as a function of spatial, seasonal, and historical fire activity predictors:
$$
\begin{aligned}
log(\frac{P(y_{i,t+1}=1)}{1-P(y_{i,t+1}=1)}) =&\beta_0 + \beta_1 \cdot \text{lat_bin}_{i} + \beta_2 \cdot \text{lon_bin}_{i} + \beta_3 \cdot \text{state}_{i}\\
+&\beta_4 \cdot \text{month_cos}_{t} + \beta_5 \cdot \text{month_sin}_{t}+\beta_6 \cdot \text{year}_{t}\\
+&\beta_7 \cdot \text{n_fires}_{i,t} + \beta_8 \cdot \text{fire_lag12}_{i,t}\\
+&\beta_9 \cdot \text{fire_roll3}_{i,t} + \beta_{10} \cdot \text{fire_roll12}_{i,t} \\
+&\beta_{11} \cdot \text{size_roll3}_{i,t} + \beta_{12} \cdot \text{size_roll12}_{i,t}
\end{aligned}
$$

In this framework, the log-odds of wildfire occurrence are modeled as a linear combination of the predictors. Each coefficient measures how a one-unit change in the predictor affects the log-odds of wildfire occurrence, holding other variables constant.

The main advantage of logistic regression is its interpretability, as the estimated coefficients provide insight into how different predictors influence wildfire risk. However, the linear structure of the model may limit its ability to capture complex nonlinear relationships and interactions among predictors.


## 2 XGBoost Model

XGBoost (Extreme Gradient Boosting) is a tree-based ensemble learning method that builds a sequence of decision trees to improve predictive performance. Each tree is trained to correct the prediction errors of the previous trees using gradient boosting.

Compared with logistic regression, XGBoost naturally captures nonlinear relationships and interactions among predictors, making it well suited for complex spatio-temporal prediction tasks such as wildfire occurrence.

Another advantage of XGBoost is its ability to handle class imbalance through parameters such as scale_pos_weight, which increases the importance of rare wildfire events during training.

While XGBoost often achieves higher predictive accuracy, it is generally less interpretable than logistic regression.




# 5 Model Fitting and Validation


```{r}

model_df <- panel %>%
  mutate(y_next = factor(y_next, levels = c(0, 1))) %>%
  select(y_next, lat_bin, lon_bin, state, 
         month_cos, month_sin, year, 
         n_fires, fire_lag12, fire_roll3, fire_roll12, size_roll3, size_roll12)
glimpse(model_df)
```




Standard k-fold cross-validation is not appropriate for time-series data because it assumes that observations are independent and identically distributed. Randomly splitting the data may allow information from future periods to influence predictions of earlier periods, resulting in temporal data leakage.

To address this issue, we implement a time-aware expanding-window cross-validation strategy. In this approach, models are trained using historical data up to a certain year and then evaluated on the following year. After each evaluation step, the training window expands to include the newly observed data.

For each validation year, we compute the Area Under the Receiver Operating Characteristic Curve (AUC) for both models. The AUC measures the ability of a model to distinguish between wildfire and non-wildfire outcomes across all possible classification thresholds. This metric is particularly suitable for imbalanced classification problems such as wildfire prediction.


```{r}

years_sorted <- sort(unique(model_df$year))
min_train_years <- 3


feat_formula <- y_next ~ lat_bin + lon_bin + state + month_cos + month_sin + year +
  n_fires + fire_lag12 + fire_roll3 + fire_roll12 + size_roll3 + size_roll12


fold_results <- tibble(
    train_end_year = years_sorted[(min_train_years):(length(years_sorted) - 2)],
    valid_year = years_sorted[(min_train_years + 1):(length(years_sorted) - 1)]
) %>%
  mutate(
      # logistic regression
      auc_glm = map2_dbl(train_end_year, valid_year, function(tr_end, va_year) {
          train <- model_df %>% filter(year <= tr_end)
          valid <- model_df %>% filter(year == va_year)
          
          # model fitting
          fit <- glm(feat_formula, data = train, family = binomial(link = "logit"))
          # prediction
          p_hat <- predict(fit, newdata = valid, type = "response")
          
          # calculating AUC
          roc_obj <- roc(response = valid$y_next, predictor = p_hat, quiet = TRUE)
          as.numeric(auc(roc_obj))}),
      
      # XGBoost
      auc_xgb = map2_dbl(train_end_year, valid_year, function(tr_end, va_year) {
          train <- model_df %>% filter(year <= tr_end)
          valid <- model_df %>% filter(year == va_year)
          
           
          train_matrix <- model.matrix(feat_formula, data = train)[, -1]  
          valid_matrix <- model.matrix(feat_formula, data = valid)[, -1]
          
          
          dtrain <- xgb.DMatrix(train_matrix, label = as.integer(as.character(train$y_next)))
          dvalid <- xgb.DMatrix(valid_matrix, label = as.integer(as.character(valid$y_next)))
          
          # 
         ratio <- sum(train$y_next == 0) / sum(train$y_next == 1)

params <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = 6,           
    eta = 0.05,              
    scale_pos_weight = ratio, 
    subsample = 0.8,
    colsample_bytree = 0.8,
    gamma = 1                
  
)




          
          fit_xgb <- xgb.train(
              params = params,
              data = dtrain,
              nrounds = 100,
              early_stopping_rounds = 10,
              watchlist = list(valid = dvalid),
              verbose = 0  
          )
          
          
          
    
          
          
          
          # prediction 
          p_hat_xgb <- predict(fit_xgb, dvalid)
          
          # AUC
          roc_obj_xgb <- roc(response = valid$y_next, predictor = p_hat_xgb, quiet = TRUE)
          as.numeric(auc(roc_obj_xgb))
      }),
      
)


print(fold_results)
```



The AUC performance of both models is visualized for each validation year.

```{r}

fold_long <- fold_results %>%
  pivot_longer(
    cols = c(auc_glm, auc_xgb),
    names_to = "model",
    values_to = "auc"
  ) %>%
  mutate(model = recode(model, auc_glm = "Logistic Regression", auc_xgb = "XGBoost"))

ggplot(fold_long, aes(x = valid_year, y = auc, color = model)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("blue", "red")) +
  labs(
    title = "Time-aware cross-validation AUC by validation year",
    x = "Validation year",
    y = "AUC",
    color = "Model"
  ) + 
  theme_minimal()


auc_glm_mean <- mean(fold_results$auc_glm)
auc_xgb_mean <- mean(fold_results$auc_xgb)
improvement <- round((auc_xgb_mean / auc_glm_mean - 1) * 100, 2)

cat('Overall, XGboost outperforms logistic regression in AUC by ', improvement, "%", sep= '')



```







The results show that XGBoost consistently outperforms logistic regression across all validation years. On average, XGBoost improves AUC by approximately $12.32\%$ compared with the logistic regression model.

This improvement indicates that the nonlinear structure of XGBoost is better able to capture complex spatial and temporal relationships in wildfire occurrence. In particular, interactions between geographic location, seasonal patterns, and historical fire activity appear to play an important role in predicting wildfire risk.





# 6 Summary

This project develops a predictive framework for forecasting notable wildfire occurrence across the United States using a spatial–temporal panel dataset. The raw wildfire incident data are transformed into a structured dataset defined at the 
$1^\circ \times 1^\circ$ grid-cell level with monthly resolution. Predictive features capturing spatial location, seasonal patterns, and historical wildfire activity are constructed while carefully avoiding temporal data leakage. Two predictive models are evaluated: logistic regression as an interpretable statistical baseline and XGBoost as a nonlinear machine learning model. Model performance is assessed using time-aware expanding-window cross-validation with AUC as the evaluation metric.
The results demonstrate that XGBoost consistently outperforms logistic regression, highlighting the importance of nonlinear relationships and interactions in wildfire prediction. Future work could further improve predictive performance by incorporating additional environmental variables such as weather conditions, vegetation indices, drought indicators, and wind patterns. These extensions could support the development of more accurate wildfire early-warning systems and help improve wildfire risk management strategies.